---
title: "Assignment 2 STAT 545"
author: "Federico Andrade-Rivas"
date: "February 5, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
``` {r libraries}
library(tidyverse)
library(gapminder)
```

## Assignment 2
####1.1 Use filter() to subset the gapminder data to three countries of your choice in the 1970’s.
```{r filter}
mi_gap <- gapminder

mi_gap %>% filter(country == "Colombia"| country == "Venezuela"| country == "Ecuador", year >= 1970 & year <= 1978)
mi_gap %>% filter(country %in% c("Colombia", "Venezuela", "Ecuador"), year %in% c(1972,1977))

```
####1.2 Use filter() to subset the gapminder data to three countries of your choice in the 1970’s.
```{r filter2}

mi_gap %>% filter(country %in% c("Colombia", "Venezuela", "Ecuador"), year %in% c(1972,1977)) %>% 
  select(country, gdpPercap)

```

####1.3 Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that’s the increase in life expectancy in your tibble. Hint: you might find the lag() or diff() functions useful.

``` {r lifeExpDrop}
  mi_gap %>% select(continent, country, year, lifeExp) %>% 
  group_by(country) %>%  
  mutate(drop = lifeExp- lag(lifeExp)) %>% 
   filter(drop < 0)

#Not sure how to use it using diff()
```
####1.4a Filter gapminder so that it shows the max GDP per capita experienced by each country. Hint: you might find the max() function useful here

```{r maxGDP}
mi_gap %>% 
  group_by(country) %>% 
  summarize(maxGDP = max(gdpPercap))
 #Not sure how to keep the year!
```

####1.4b Filter gapminder to contain six rows: the rows with the three largest GDP per capita, and the rows with the three smallest GDP per capita. Be sure to not create any intermediate objects when doing this (with, for example, the assignment operator). Hint: you might find the sort() function useful, or perhaps even the dplyr::slice() function.

```{r 6rows}
#mi_gap %>% select(country, year, gdpPercap) %>% 
 # group_by(country) %>% 
  #summarize(sort(gdpPercap))

```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
