---
title: "Assignment 2 STAT 545"
author: "Federico Andrade-Rivas"
date: "February 5, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
``` {r libraries}
library(tidyverse)
library(gapminder)
```

## Assignment 2
####1.1 Use filter() to subset the gapminder data to three countries of your choice in the 1970’s.
```{r filter}
mi_gap <- gapminder

mi_gap %>% filter(country == "Colombia"| country == "Venezuela"| country == "Ecuador", year >= 1970 & year <= 1978)
mi_gap %>% filter(country %in% c("Colombia", "Venezuela", "Ecuador"), year %in% c(1972,1977))

```
####1.2 Use filter() to subset the gapminder data to three countries of your choice in the 1970’s.
```{r filter2}

mi_gap %>% filter(country %in% c("Colombia", "Venezuela", "Ecuador"), year %in% c(1972,1977)) %>% 
  select(country, gdpPercap)

```

####1.3 Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that’s the increase in life expectancy in your tibble. Hint: you might find the lag() or diff() functions useful.

``` {r lifeExpDrop}
  mi_gap %>% select(continent, country, year, lifeExp) %>% 
  group_by(country) %>%  
  mutate(drop = lifeExp- lag(lifeExp)) %>% 
   filter(drop < 0) 

```




## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
