---
title: "Chapter 17"
author: "Federico Andrade-Rivas"
date: "April 21, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### 17.1 Vectors are everywhere
Square brackets are used for isolating elements of a vector for inspection, modification, etc. This is often called indexing. BTW, indexing begins at 1 in R, unlike many other languages that index from 0.

```{r vector}

 x <- 3 * 4
x

is.vector(x)
length(x)

x[2] <- 100
x

(x[5] <- 3)
x

x[11]

x[0]

```

NOTE: R is built to work with vectors. Many operations are vectorized, i.e. by default they will happen component-wise when given a vector as input. Novices often don’t internalize or exploit this and they write lots of unnecessary for loops.


```{r}
n <- 8
set.seed(1)
(w <- round(rnorm(n), 2)) # numeric floating point
#> [1] -0.63  0.18 -0.84  1.60  0.33 -0.82  0.49  0.74
(x <- 1:n) # numeric integer
#> [1] 1 2 3 4 5 6 7 8
## another way to accomplish by hand is x <- c(1, 2, 3, 4, 5, 6, 7, 8)
(y <- LETTERS[1:n]) # character
#> [1] "A" "B" "C" "D" "E" "F" "G" "H"
(z <- runif(n) > 0.3) # logical
#> [1]  TRUE  TRUE  TRUE  TRUE  TRUE FALSE  TRUE FALSE

str(w)
#>  num [1:8] -0.63 0.18 -0.84 1.6 0.33 -0.82 0.49 0.74
length(x)
#> [1] 8
is.logical(y)
#> [1] FALSE
as.numeric(z)
#> [1] 1 1 1 1 1 0 1 0
```


####17.2 Indexing a vector

We’ve said, and even seen, that square brackets are used to index a vector. There is great flexibility in what one can put inside the square brackets and it’s worth understanding the many options. They are all useful, just in different contexts.

Most common, useful ways to index a vector:

* logical vector: keep elements associated with TRUE’s, ditch the FALSE’s
* vector of positive integers: specifying the keepers
* vector of negative integers: specifying the losers
* character vector: naming the keepers




```{r}
w
#> [1] -0.63  0.18 -0.84  1.60  0.33 -0.82  0.49  0.74
names(w) <- letters[seq_along(w)]
w
#>     a     b     c     d     e     f     g     h 
#> -0.63  0.18 -0.84  1.60  0.33 -0.82  0.49  0.74
w < 0
#>     a     b     c     d     e     f     g     h 
#>  TRUE FALSE  TRUE FALSE FALSE  TRUE FALSE FALSE
which(w < 0)
#> a c f 
#> 1 3 6
w[w < 0]
#>     a     c     f 
#> -0.63 -0.84 -0.82
seq(from = 1, to = length(w), by = 2)
#> [1] 1 3 5 7
w[seq(from = 1, to = length(w), by = 2)]
#>     a     c     e     g 
#> -0.63 -0.84  0.33  0.49
w[-c(2, 5)]
#>     a     c     d     f     g     h 
#> -0.63 -0.84  1.60 -0.82  0.49  0.74
w[c('c', 'a', 'f')]
#>     c     a     f 
#> -0.84 -0.63 -0.82

```

#### 
17.4 Creating a data.frame explicitly

In data analysis, we often import data into data.frame via read.table(). But one can also construct a data.frame directly using data.frame().

```{r}

n <- 8
set.seed(1)
(jDat <- data.frame(w = round(rnorm(n), 2),
                    x = 1:n,
                    y = I(LETTERS[1:n]),
                    z = runif(n) > 0.3,
                    v = rep(LETTERS[9:12], each = 2)))

str(jDat)

mode(jDat)

class(jDat)

```


```{r}
is.list(jDat) # data.frames are lists

jDat[[5]] # this works but I prefer ...

jDat$v # using dollar sign and name, when possible

jDat[c("x", "z")] # get multiple variables


str(jDat[c("x", "z")]) # returns a data.frame

identical(subset(jDat, select = c(x, z)), jDat[c("x", "z")])

```

#### 17.6 Creating arrays, e.g. matrices

There are three main ways to create a matrix. It goes without saying that the inputs must comply with the requirement that all matrix elements are the same flavor. If that’s not true, you risk an error or, worse, silent conversion to character.

  *  Filling a matrix with a vector
  * Glueing vectors together as rows or columns
  *  Conversion of a data.frame

Let’s demonstrate. Here we fill a matrix with a vector, explore filling by rows and giving row and columns at creation. Notice that recycling happens here too, so if the input vector is not large enough, R will recycle it.

```{r}
matrix(1:15, nrow = 5)

matrix("hola!", nrow = 3, ncol = 6)

matrix(c("hola!", "tu?"), nrow = 3, ncol =6 )

matrix(1:15, nrow = 5, byrow = TRUE)

matrix(1:15, nrow = 5,
       dimnames = list(paste0("row", 1:5),
                       paste0("col", 1:3)))

```



Here we create a matrix by glueing vectors together. Watch the vector names propagate as row or column names.

```{r}
vec1 <- 5:1
vec2 <- 2^(1:5)
cbind(vec1, vec2)

rbind(vec1, vec2)

```

Here we create a matrix from a data.frame.

```{r}
vecDat <- data.frame(vec1 = 5:1,
                     vec2 = 2^(1:5))
str(vecDat)

vecMat <- as.matrix(vecDat)
str(vecMat)

```

Here we create a matrix from a data.frame, but experience the “silently convert everything to character” fail. As an added bonus, I’m also allowing the “convert character to factor” thing to happen when we create the data.frame initially. Let this be a reminder to take control of your objects!

```{r}

multiDat <- data.frame(vec1 = 5:1,
                       vec2 = paste0("hi", 1:5))
str(multiDat)

(multiMat <- as.matrix(multiDat))

str(multiMat)

```



