---
title: "Chapter7_Fede"
author: "Federico Andrade-Rivas"
date: "January 29, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Loading Libraries

```{r library}
library(tidyverse)
library(gapminder)
```


##Create a copy of gapminder

``` {r copyGap}
library(gapminder)
my_gap <-gapminder
my_gap %>% filter(country == "Canada")  

#Storing the output as an object

my_precious <- my_gap %>% filter(country =="Canada")

```

``` {r Mutate}
 my_gap %>% mutate(gdp = pop*gdpPercap)

# Create a new variable that is gdpPercap divided by Canadian gdpPercap, taking care that I always divide two numbers that pertain to the same year

cgdp <- my_gap %>% filter(country =="Canada")

my_gap <- my_gap %>% mutate(temp = rep(cgdp$gdpPercap, nlevels(country)),
                            gdpRelaC = gdpPercap/temp,
                            temp = NULL
                            )
#Note: nlevels is used to create a variable that is as long as the number of levels per country. All countries have data for the same years. Otherwise this would not work 

my_gap %>%  filter(country == "Canada") %>% select(country, year, gdpPercap, gdpRelaC)

#Distribution of gdp related with the Canadian one

summary(my_gap$gdpRelaC)


```
#!!!!Remember: Trust No One. Including (especially?) yourself. Always try to find a way to check that you’ve done what meant to. Prepare to be horrified.

##Arrange function
```{r arrange}
#Use arrange() to row-order data in a principled way
#Example to arrange by year then country
my_gap %>% arrange(year, country)

#Example to arrange by continent then country
my_gap %>% arrange(continent, country)

#example one year but sorted by life expectancy
 my_gap %>% filter(year == 1952) %>% arrange(gdpPercap) %>% select(country, continent, year, gdpPercap)

 #NOw in descending order for 2007
 
 my_gap %>% filter(year == 2007) %>% arrange(desc(gdpPercap)) %>% select(country, continent, year, gdpPercap)

 #"I advise that your analyses NEVER rely on rows or variables being in a specific order. But it’s still true that human beings write the code and the interactive development process can be much nicer if you reorder the rows of your data as you go along. Also, once you are preparing tables for human eyeballs, it is imperative that you step up and take control of row order."
```

##Rename, Select, and Group by

```{r Ch7rename}
#camelCase Vs snake_case
#Rename variables (not assigning!)

my_gap %>% rename(life_exp = lifeExp,
                  gdp_percap = gdpPercap,
                  gdp_percap_rel = gdpRelaC)

#select can rename and reposition 
#select() can rename the variables you request to keep.
#select() can be used with everything() to hoist a variable up to the front of the tibble.

my_gap %>% 
  filter(country == "Burundi", year > 1996) %>% 
  select(yr = year, life_exp = lifeExp, gdp_Percap = gdpPercap) %>% 
  select(gdp_Percap, everything())

#Counting things up

my_gap %>% 
  group_by(continent) %>% 
  summarize(n = n())

#Same info as R base table() but issues with computing

table(my_gap$continent)

#tally() counts rows

my_gap %>% 
  group_by(continent) %>% 
  tally()

#The count() function is an even more convenient function that does both grouping and counting.

my_gap %>% 
  count(continent) 

#Use the n_distinct() function to count the number of distinct countries within each continent.

my_gap %>% 
  group_by(continent) %>% 
  summarize(n = n(),
            n_countries = n_distinct(country))


```

###TO DO 7.8.2 General summarization


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
