---
title: "Chapter18"
author: "Federico Andrade-Rivas"
date: "April 22, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####18.2 Load the Gapminder data

```{r}
library(gapminder)
str(gapminder)

```

Say you’ve got a numeric vector, and you want to compute the difference between its max and min. lifeExp or pop or gdpPercap are great examples of a typical input. You can imagine wanting to get this statistic after we slice up the Gapminder data by year, country, continent, or combinations thereof.


```{r}
min(gapminder$lifeExp)
max(gapminder$lifeExp)
range(gapminder$lifeExp)

## some natural solutions
max(gapminder$lifeExp) - min(gapminder$lifeExp)

with(gapminder, max(lifeExp) - min(lifeExp))

range(gapminder$lifeExp)[2] - range(gapminder$lifeExp)[1]

with(gapminder, range(lifeExp)[2] - range(lifeExp)[1])

diff(range(gapminder$lifeExp))


```

####18.5 Turn the working interactive code into a function

Add NO new functionality! Just write your very first R function.

```{r}
max_minus_min <- function(x) max(x) - min(x)

max_minus_min(gapminder$lifeExp)
```

####18.6 Test your function
#####18.6.1 Test on new inputs

Pick some new artificial inputs where you know (at least approximately) what your function should return.

```{r}
max_minus_min(1:22)

max_minus_min(runif(1000))
```

#####
18.6.2 Test on real data but different real data

```{r}
max_minus_min(gapminder$pop)

max_minus_min(gapminder$gdpPercap)

```

#### 18.6.3 Test on weird stuff

Now we try to break our function. Don’t get truly diabolical (yet). Just make the kind of mistakes you can imagine making at 2am when, 3 years from now, you rediscover this useful function you wrote. Give your function inputs it’s not expecting.


```{r}

#max_minus_min <- function(x) max(x) - min(x)

#max_minus_min(gapminder)

#max_minus_min(gapminder$continent)

#max_minus_min("hola")
```

##### 18.6.4 I will scare you now


```{r}
max_minus_min(gapminder[c('lifeExp', 'gdpPercap', 'pop')])

max_minus_min(c(TRUE, TRUE, FALSE, TRUE, TRUE))

```
####18.7 Check the validity of arguments

For functions that will be used again – which is not all of them! – it is good to check the validity of arguments. This implements a rule from the Unix philosophy:

   * Rule of Repair: When you must fail, fail noisily and as soon as possible.


#### 18.7.1 stop if not

stopifnot() is the entry level solution. I use it here to make sure the input x is a numeric vector.

```{r}

mmm <- function(x) {
  stopifnot(is.numeric(x))
  max(x) - min(x)
}

#mmm(gapminder)

#mmm(gapminder$country)

#mmm(gapminder[c('lifeExp', 'gdpPercap', 'pop')])

#mmm(c(TRUE, TRUE, FALSE, TRUE, TRUE))



```

####18.7.2 if then stop

stopifnot() doesn’t provide a very good error message. The next approach is very widely used. Put your validity check inside an if() statement and call stop() yourself, with a custom error message, in the body.

```{r}
mmm2 <- function(x) {
  if(!is.numeric(x)) {
    stop("Lo siento, esta función sólo funciona con valores numéricos\n", "Usted a entrado un objeto de clase: :", class(x)[1])
  }
    max(x) - min(x)
}

##mmm2(gapminder)
```

In addition to a gratuitous apology, the error raised also contains two more pieces of helpful info:

*    Which function threw the error.
*   Hints on how to fix things: expected class of input vs actual class.

If it is easy to do so, I highly recommend this template: “you gave me THIS, but I need THAT”.

The tidyverse style guide has a very useful chapter on how to construct error messages.

