---
title: "Chapter 13"
author: "Federico Andrade-Rivas"
date: "March 19, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Chapter 13 Dates and Times

```{r Loadpackages}
library(tidyverse)
library(lubridate)
library(nycflights13)
```


Get times! Lubridate, and Rbase
```{r times}
today()
#lubridate

Sys.Date()
#Base R

str(Sys.Date())

class(Sys.Date())

Sys.time()
#Rbase

now()
#Lubridate

##They both give you something of class POSIXct in R jargon.
str(Sys.time())

class(Sys.time())

str(now())

class(now())

```

###13.5 Get date or date-time from character

An approach is to use the helpers provided by lubridate. They automatically work out the format once you specify the order of the component. To use them, identify the order in which year, month, and day appear in your dates, then arrange “y”, “m”, and “d” in the same order. That gives you the name of the 
lubridate function that will parse your date. For example:

These functions also take unquoted numbers. This is the most concise way to create a single date/time object, as you might need when filtering date/time data. ymd() is short and unambiguous.

``` {r fromStrings}
ymd("2017-01-31")

mdy("January 31st, 2017")

dmy("2017-01-31")

ymd(20170131)

```
To create a date-time, add an underscore and one or more of “h”, “m”, and “s” to the name of the parsing function:

```{r timehour}
ymd_hms("2017-01-31 21:34:58")

mdy_hm("01/31/2017 08:01")

##You can also force the creation of a date-time from a date by supplying a timezone:
ymd(20170131, tz = "UTC")

```

###13.6 Build date or date-time from parts

Instead of a single string, sometimes you’ll have the individual components of the date-time spread across multiple columns. This is what we have in the flights data:

```{r diffColumns}

flights %>% select(year, month, day, hour, minute)
```

To create a date/time from this sort of input, use make_date() for dates, or make_datetime() for date-times:

```{r dateTime}
flights %>% select(year, month, day, hour, minute) %>% 
            mutate(departure = make_datetime(year, month, day, hour, minute))

```

Let’s do the same thing for each of the four time columns in flights. The times are represented in a slightly odd format, so we use modulus arithmetic to pull out the hour and minute components. Once I’ve created the date-time variables, I focus in on the variables we’ll explore in the rest of the chapter.

```{r maketime}
make_datetime_100 <- function(year, month, day, time) {
make_datetime(year, month, day, time %/% 100, time %% 100)}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))
##NOTE the ends_with() function. Selects the variable has the string provided at the end =)

flights_dt


```

Plotting distribution of departure times across the year
```{r, echo=FALSE }
flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day

```

Or within a single day:
```{r singleday, echo=FALSE}
flights_dt %>% 
  filter(dep_time < ymd(20130102)) %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes

```
Note that when you use date-times in a numeric context (like in a histogram), 1 means 1 second, so a binwidth of 86400 means one day. For dates, 1 means 1 day.

###16.2.3 From other types (R Studio Book)

You may want to switch between a date-time and a date. That’s the job of as_datetime() and as_date():

Sometimes you’ll get date/times as numeric offsets from the “Unix Epoch”, 1970-01-01. If the offset is in seconds, use as_datetime(); if it’s in days, use as_date().


```{r}
as_datetime(today())
#> [1] "2020-01-15 UTC"
as_date(now())
#> [1] "2020-01-15"


as_datetime(60 * 60 * 10)
#> [1] "1970-01-01 10:00:00 UTC"
as_date(365 * 10 + 2)
#> [1] "1980-01-01"

```

###16.2.4 Exercises
1. What happens if contains invalid dates?

```{r}
ymd(c("2010-10-10", "bananas"))


```
2. What does the tzone argument to today() do? Why is it important? 
It specifies the time zone. The default is the computer's time zone

3. Use the appropriate lubridate function to parse each of the following dates:
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014

```{r}
d1 <- "January 1, 2010"
mdy(d1)

d2 <- "2015-Mar-07"
ymd(d2)

d3 <- "06-Jun-2017"
dmy(d3)

d4 <- c("August 19 (2015)", "July 1 (2015)")
mdy(d4)

d5 <- "12/30/14" # Dec 30, 2014
mdy(d5)

```

####16.3.1 Getting components

You can pull out individual parts of the date with the accessor functions year(), month(), mday() (day of the month), yday() (day of the year), wday() (day of the week), hour(), minute(), and second().

```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)

month(datetime)

mday(datetime)
#mday <- day of the month

yday(datetime)
# yday <- day of the year

wday(datetime)
#wday <- day of the week 

hour(datetime)
minute(datetime)
second(datetime)

```

For month() and wday() you can set label = TRUE to return the abbreviated name of the month or day of the week. Set abbr = FALSE to return the full name.

```{r}

month(datetime, label =TRUE, abbr = TRUE)
month(datetime, label =TRUE, abbr = FALSE)

wday(datetime, label = TRUE, abbr = TRUE)
wday(datetime, label = TRUE, abbr = FALSE)

```

We can use wday() to see that more flights depart during the week than on the weekend:
```{r}
flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) +
    geom_bar()
```

There’s an interesting pattern if we look at the average departure delay by minute within the hour. It looks like flights leaving in minutes 20-30 and 50-60 have much lower delays than the rest of the hour!

```{r}
flights_dt %>% 
  mutate(minute = minute(dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, avg_delay)) +
    geom_line()
```

Interestingly, if we look at the scheduled departure time we don’t see such a strong pattern:
```{r}
sched_dep <- flights_dt %>% 
  mutate(minute = minute(sched_dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n())

ggplot(sched_dep, aes(minute, avg_delay)) +
  geom_line()
```
 
So why do we see that pattern with the actual departure times? Well, like much data collected by humans, there’s a strong bias towards flights leaving at “nice” departure times. Always be alert for this sort of pattern whenever you work with data that involves human judgement!

```{r}
ggplot(sched_dep, aes(minute, n)) +
  geom_line()
```

####16.3.2 Rounding

An alternative approach to plotting individual components is to round the date to a nearby unit of time, with floor_date(), round_date(), and ceiling_date(). Each function takes a vector of dates to adjust and then the name of the unit round down (floor), round up (ceiling), or round to. This, for example, allows us to plot the number of flights per week:

```{r}
flights_dt %>% 
  count(week = floor_date(dep_time, "week")) %>% 
  ggplot(aes(week, n)) +
    geom_line()
```

####16.3.3 Setting components

You can also use each accessor function to set the components of a date/time:

```{r}
(datetime <- ymd_hms("2016-07-08 12:34:56"))
#> [1] "2016-07-08 12:34:56 UTC"

year(datetime) <- 2020
datetime
#> [1] "2020-07-08 12:34:56 UTC"
month(datetime) <- 01
datetime
#> [1] "2020-01-08 12:34:56 UTC"
hour(datetime) <- hour(datetime) + 1
datetime
#> [1] "2020-01-08 13:34:56 UTC"
```

Alternatively, rather than modifying in place, you can create a new date-time with update(). This also allows you to set multiple values at once.

```{r}
update(datetime, year = 2020, month = 2, mday = 2, hour = 2)
#> [1] "2020-02-02 02:34:56 UTC"
```

If values are too big, they will roll-over:
```{r}
ymd("2015-02-01") %>% 
  update(mday = 30)
#> [1] "2015-03-02"
ymd("2015-02-01") %>% 
  update(hour = 400)
#> [1] "2015-02-17 16:00:00 UTC"
```

You can use update() to show the distribution of flights across the course of the day for every day of the year:

```{r}
flights_dt %>% 
  mutate(dep_hour = update(dep_time, yday = 300)) %>% 
  ggplot(aes(dep_hour)) +
    geom_freqpoly(binwidth = 300)
```

###16.3.4 Exercises TO DO ONCE I TAKE GGPLOT (GOOD EXERCISE)


1 How does the distribution of flight times within a day change over the course of the year?

2 Compare dep_time, sched_dep_time and dep_delay. Are they consistent? Explain your findings.

3 Compare air_time with the duration between the departure and arrival. Explain your findings. (Hint: consider the location of the airport.)

4 How does the average delay time change over the course of a day? Should you use dep_time or sched_dep_time? Why?

5 On what day of the week should you leave if you want to minimise the chance of a delay?

6 What makes the distribution of diamonds$carat 
and 
flights$sched_dep_time similar?

7 Confirm my hypothesis that the early departures of flights in minutes 20-30 and 50-60 are caused by scheduled flights that leave early. Hint: create a binary variable that tells you whether or not a flight was delayed.

### 16.4 Time spans

Next you’ll learn about how arithmetic with dates works, including subtraction, addition, and division. Along the way, you’ll learn about three important classes that represent time spans:

*durations, which represent an exact number of seconds.
*periods, which represent human units like weeks and months.
*intervals, which represent a starting and ending point.

In R, when you subtract two dates, you get a difftime object:


```{r}

# How old is Hadley?
h_age <- today() - ymd(19791014)
h_age
#> Time difference of 14703 days

```

A difftime class object records a time span of seconds, minutes, hours, days, or weeks. This ambiguity can make difftimes a little painful to work with, so lubridate provides an alternative which always uses seconds: the duration.

```{r}
as.duration(h_age)
#> [1] "1270339200s (~40.25 years)"
```


