---
title: "Chapter 10 STAT545"
author: "Federico Andrade-Rivas"
date: "March 4, 2020"
output: github_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## FACTORS


```{r LoadLibraries, include= FALSE}
library(tidyverse)
library(gapminder)
```

###10.4 Factor Inspection

``` {r factorInspection}

str(gapminder$continent)

levels(gapminder$continent)

nlevels(gapminder$continent)

class(gapminder$continent)

fct_count(gapminder$continent)

gapminder %>% count(continent)

```



###10.5 Dropping unused levels 


```{r dropping_leveslintro}
nlevels(gapminder$country)

#Watch what happens to the levels of country (= nothing) when we filter Gapminder to a handful of countries.

h_countries <- c("Colombia", "Canada", "South Africa", "Ecuador", "Costa Rica", "Peru")
h_gap <- gapminder %>% filter(country %in% h_countries)
nlevels(h_gap$country)

#Even though h_gap only has data for a handful of countries, we are still schlepping around all 142 levels from the original gapminder tibble.
```

Dropping levels using different strategies
```{r dropping_levels}
h_gapdropped <- h_gap %>% droplevels()
nlevels(h_gapdropped$country)

h_gap$country %>% fct_drop() %>% 
                  levels()

```

### Exercise 10.5. Filter the gapminder data down to rows where population is less than a quarter of a million, i.e. 250,000. Get rid of the unused factor levels for country and continent in different ways, such as:

```{r exercise10.5}

nlevels(gapminder$country)
nlevels(gapminder$continent)

#using droplevels()
gap_Lessdrop <- gapminder %>% filter(pop < 250000) %>% 
                              droplevels()  
nlevels(gap_Lessdrop$country)
nlevels(gap_Lessdrop$continent)            
 
# using fct_drop() inside mutate() NOT WORKING

#gap_lessfct <- gapminder %>% mutate(countryless1qrt = pop < 250000, fct_drop(country)) %>% 
 #filter(countryless1qrt == TRUE) #%>% 
   #nlevels(gap_lessfct$country)

```


### Change order of the levels, principled

Order by frequency forward and backwards
``` {r changelevelsorder}

gapminder$continent %>% levels() 


gapminder$continent %>% fct_infreq() %>% levels()


gapminder$continent %>% fct_infreq() %>% fct_rev() %>% levels()

```

Order country by other factor (The factor is the grouping variable and the default summarizing function is median() but you can specify something else.). Usse .desc for backwards
``` {r sorder}

fct_reorder(gapminder$country, gapminder$lifeExp) %>% levels() %>% head() 

fct_reorder(gapminder$country, gapminder$lifeExp, min) %>% levels() %>% head()

fct_reorder(gapminder$country, gapminder$lifeExp, median, .desc = TRUE) %>% levels() %>% head()
  
```

```{r example plot, echo=FALSE}
gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia")
ggplot(gap_asia_2007, aes(x = lifeExp, y = country)) + geom_point()

ggplot(gap_asia_2007, aes(x = lifeExp, y = fct_reorder(country, lifeExp))) +
  geom_point()

```

Use fct_reorder2() when you have a line chart of a quantitative x against another quantitative y and your factor provides the color. This way the legend appears in some order as the data! Contrast the legend on the left with the one on the right.

```{r exampleplot, echo=FALSE}
h_countries <- c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela")
h_gap <- gapminder %>%
  filter(country %in% h_countries) %>% 
  droplevels()

ggplot(h_gap, aes(x = year, y = lifeExp, color = country)) +
  geom_line()

ggplot(h_gap, aes(x = year, y = lifeExp,
                  color = fct_reorder2(country, year, lifeExp))) +
  geom_line() +
  labs(color = "country")

```

Change order of the levels, “because I said so”

```{r changelevelsbecause}

h_gap$country %>% levels()

h_gap$country %>% fct_relevel("Thailand", "Venezuela", "Romania") %>% levels()
 
```

Changing names of levels

```{r changenames}

i_gap <- gapminder %>% filter(country %in% c("Colombia", "Ecuador", "Chile", "Mexico", "Argentina")) %>% droplevels()

i_gap$country %>% fct_recode("Chimbchombianos" = "Colombia", "Merito" = "Mexico") %>% levels()

```

Exercise: Isolate the data for "Australia", "Korea, Dem. Rep.", and "Korea, Rep." in the 2000x. Revalue the country factor levels to "Oz", "North Korea", and "South Korea".

```{r exerciselevels}

Revalue_gap <- gapminder %>% filter(country %in% c("Australia", "Korea, Dem. Rep.", "Korea, Rep."), year >= 2000) %>% droplevels()

Revalue_gap$country %>% fct_recode("Oz" = "Australia", "North Korea" = "Korea, Dem. Rep.", "South Korea" = "Korea, Rep.") %>% levels()
Revalue_gap

```

Let’s create two data frames, each with data from two countries, dropping unused factor levels.

```{r gRowaFactor}

df1 <- gapminder %>% filter(country %in% c("Colombia", "South Africa"), year >= 2000) %>%
                    droplevels()

df2 <- gapminder %>% filter(country %in% c("Canada", "Peru"), year  >= 2000) %>% 
                      droplevels()

nlevels(df1$country)
nlevels(df2$country)


```
Combining the levels using fct_c(). And exploring different forms of row binding

```{r combining}
fct_c(df1$country, df2$country)

bind_rows(df1, df2)

rbind(df1, df2)

```




