---
title: "Chapter 10 STAT545"
author: "Federico Andrade-Rivas"
date: "March 4, 2020"
output: github_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## FACTORS


```{r cars, include= FALSE}
library(tidyverse)
library(gapminder)
```

###10.4 Factor Inspection

``` {r factorInspection}

str(gapminder$continent)

levels(gapminder$continent)

nlevels(gapminder$continent)

class(gapminder$continent)

fct_count(gapminder$continent)

gapminder %>% count(continent)

```



###10.5 Dropping unused levels 


```{r dropping_leveslintro}
nlevels(gapminder$country)

#Watch what happens to the levels of country (= nothing) when we filter Gapminder to a handful of countries.

h_countries <- c("Colombia", "Canada", "South Africa", "Ecuador", "Costa Rica", "Peru")
h_gap <- gapminder %>% filter(country %in% h_countries)
nlevels(h_gap$country)

#Even though h_gap only has data for a handful of countries, we are still schlepping around all 142 levels from the original gapminder tibble.
```

Dropping levels using different strategies
```{r dropping_levels}
h_gapdropped <- h_gap %>% droplevels()
nlevels(h_gapdropped$country)

h_gap$country %>% fct_drop() %>% 
                  levels()

```

### Exercise 10.5. Filter the gapminder data down to rows where population is less than a quarter of a million, i.e. 250,000. Get rid of the unused factor levels for country and continent in different ways, such as:

```{r exercise10.5}

nlevels(gapminder$country)
nlevels(gapminder$continent)

#using droplevels()
gap_Lessdrop <- gapminder %>% filter(pop < 250000) %>% 
                              droplevels()  
nlevels(gap_Lessdrop$country)
nlevels(gap_Lessdrop$continent)            
 
# using fct_drop() inside mutate() NOT WORKING

#gap_lessfct <- gapminder %>% mutate(countryless1qrt = pop < 250000, fct_drop(country)) %>% 
 #  filter(countryless1qrt == TRUE) %>% 
   
#nlevels(gap_lessfct$country)

```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
