---
title: "Assignment 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(gapminder)
library(tidyverse)
```

##Exercise 1
```{r exercise 1}
# 1.1 and 1.2
knitr::kable(
gapminder %>% 
  filter(country %in% c("Canada", "Japan", "Mexico"),
       year >= 1970 & year <= 1979) %>% 
  select(country, gdpPercap)
)

# 1.3
gapminder %>%
  group_by(country) %>% 
  mutate(le_delta = lifeExp - lag(lifeExp)) %>%
  filter(le_delta < 0)

# 1.4
gapminder %>%
  group_by(country) %>% 
  summarize(max_gdpPercap = max(gdpPercap, na.rm = TRUE)) %>% 
  arrange(desc(max_gdpPercap))
  
# 1.5
gapminder %>% 
  filter(country == "Canada") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    scale_x_log10()
```

##Exercise 2
Categorical var = continent  
Quantitative var = lifeExp
```{r exercise 2, echo=FALSE}
# Country list
levels(gapminder$continent)
# Life expectancy summary stats
summary(gapminder$lifeExp)

gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(avg_le = mean(lifeExp), na.rm = TRUE) %>% 
  ggplot(aes(x = continent, y = avg_le)) +
  geom_bar(stat = "identity") +
  ggtitle("Mean life expectancy by continent in 2007")
```

##Exercise 3
```{r exercise , echo=FALSE, fig.width=8, fig.height=12.5}
# 
gapminder %>%
  ggplot(aes(x = factor(year), y = country, fill = lifeExp)) +
         geom_tile(color = "white") +
  scale_fill_distiller(palette ="RdYlBu")

```

##Recycling

> filter(gapminder, country == c("Rwanda", "Afghanistan"))  
> Not sure why this doesn't work  
> It returns half of the values for each country -- years ending in 7 for Afghanistan and years ending in 2 for Rwanda  
> It works if you pipe instead of concatinate although I don't know why  
> filter(gapminder, country == "Rwanda" | country == "Afghanistan")