---
title: "Stat545.com Practice"
output: 
  word_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 5 R Markdown

This is a practice R Markdown document and used for running chapter 5 examples.  
First, let's start by investigating the power of tidyverse and tibbles.

```{r Chunk title goes here, echo = TRUE}
library(tidyverse)
library(gapminder)
str(gapminder)
class(gapminder)
head(gapminder)
tail(gapminder)
as_tibble(iris)
names(gapminder)
```

<!-- Note that this type of commenting does NOT show up on your final knit document -->
Note that lenght command returns the number of columns NOT the number of rows
```{r, echo=TRUE}
length(gapminder)
nrow(gapminder)
ncol(gapminder)
dim(gapminder)
```

Let's make a plot with base R graphics
```{r, echo=FALSE}
plot(lifeExp ~ year, gapminder)
hist(gapminder$lifeExp)
```

Ok, why does the histogram look like it does?
```{r, echo=TRUE}
summary(gapminder$year)
table(gapminder$year)
class(gapminder$year)
```

So that was an integer.  
What about other variables in the dataset? Let's pick a categorical one.  
*note that this paragraph contains line breaks -which means TWO spaces at the end*  
*Does it look like it in the rendered doc?*  
**Still not for some reason...**
```{r,echo=TRUE}
class(gapminder$continent)
summary(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
```

```{r Barplot - graph title, echo=TRUE}
barplot(table(gapminder$continent))
```
Don't forget that R is storing those levels as integers i.e. '1', '2', '3', etc under the hood!!! This will trip up unwary programmers, so don't forget to properly care for your factors. *note that this paragraph was written without line breaks, does it make a difference to the final formatting?*


Here is a much nicer ggplot graph
```{r, echo=TRUE}
## we exploit the fact that ggplot was installed and loaded via the tidyverse
p <- ggplot(filter(gapminder, continent != "Oceania"),
            aes(x = gdpPercap, y = lifeExp)) # just initializes
p <- p + scale_x_log10() #log the x axis the right way
p + geom_point() #scatterplot
p + geom_point(aes(color = continent)) #map the continents to color
p + geom_point(alpha = (1/3), size = 3) +
  geom_smooth(lwd = 3, se = F) 
p + geom_point(alpha = (1/3), size = 3) + 
  facet_wrap(~continent) +
  geom_smooth(lwd = 3, se = F) 

```
That's a wrap for Chapter 5 and practicing with R markdown!!