---
title: "Stat 545 homework 2"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r echo = FALSE}
library(tidyverse)
library(gapminder)
library(knitr)
```

## Exercise 1 - Basic dplyr
```{r exercise 1}
# Examine the dataset to make sure it imported properly
glimpse(gapminder)
head(gapminder)
tail(gapminder)
colSums(is.na(gapminder))


# 1.1 and 1.2 Filter down to year and countries of choice
knitr::kable(
gapminder %>% 
  filter(year >= 1970 & year <= 1979, 
         country == "Canada" | country == "Colombia" | country == "Cote d'Ivoire") %>%
  select(country, gdpPercap)
)

# 1.3 Drop in life expectancy
DT::datatable(
gapminder %>% 
  group_by(country) %>% 
  mutate(delta_life_exp = gdpPercap - lag(gdpPercap)) %>%
  filter(delta_life_exp < 0) 
)
# Note that Dplyr functions automatically removes NAs and don't accept the na.rm argument

# 1.4a Max GDP per country
gapminder %>% 
  group_by(country) %>% 
  summarize(max_gdp = max(gdpPercap))

# 1.4b Three countries with the largest GPD and lowest GDP
gapminder %>% 
  group_by(country) %>% 
  summarise(max_gdp = max(gdpPercap)) %>% 
  filter(min_rank(max_gdp) <= 3 | min_rank(desc(max_gdp)) <= 3) %>% 
  arrange(max_gdp)

# 1.5 Scatterplot of Canada life expectancy vs GDP
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    scale_x_log10()
```
Apparently to use DT::datatable() you need to render the doc as an html, not as a git_hub doc

## Exercise 2 - explore individual variables
2.1 Categorical variable - continent
```{r exercise 2 - categorical}
n_distinct(gapminder$continent)
levels(gapminder$continent)
table(gapminder$continent)
ggplot(gapminder, aes(x = continent)) +
  geom_bar()
```


```{r table attempt, results='asis'}
kable(table(gapminder$continent), caption = "Frequency by continent")
```

2.2 Quantitative variable - life expectancy
```{r exercise 2 - quantitative}
# Range
range(gapminder$lifeExp)

# 5 var summary and IQR
summary(gapminder$lifeExp)
boxplot(gapminder$lifeExp)
```

## Recycling
The issue with this command is that it should produce a table of 24 rows, and instead it is only 12
```{r}
table(gapminder$country) #Sadly table() does not knit with the datatable function

t <- filter(gapminder, country == c("Rwanda", "Afghanistan"))

nrow(t)
```

I'm not really sure why this is, but based on the title of the exercise—and the fact that each year is only in the filtered object once- I'd guess that for each set of unique variables (minus country) it alternates looking for Rwanda and then Afghanistan??? 

Apparently you don't get an error if the dataframe has a number of rows equal to the value of the target (ie 2 countries, so if the df is even)

Apparently this is also related to how == works, nothing to do with dplyr 

See this answer on stacks overflow which is the closest answer I've found: https://stackoverflow.com/questions/25647470/filter-multiple-values-on-a-string-column-in-dplyr

Better options: use %in% or use multiple conditions with or (|)
```{r better than trusting recycling}
# Option 1: Use group membership
c_names <- c("Rwanda", "Afghanistan")
filter(gapminder, country %in% c_names)

# Option 2: Use 'or'
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```