---
title: "Functions in R"
author: "Michelle Coombe"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Load libraries and data
```{r}
library(tidyverse)
library(gapminder)
library(lubridate)
```
Notes in this document are primarily from the original [stat545](https://stat545.com/functions-part1.html), but also from the R for Data Science [ebook](https://r4ds.had.co.nz/functions.html), and the newer [ubcstat545](https://stat545guidebook.netlify.com/functional-programming-in-r-part-i.html).

## Random notes to help understand how R works with functions

### Indexing and vectorized operations
Using square brackets to isolate elements of a vector = indexing. Index value begin at '1' in R, which seems obvious, but apparently most other languages start at '0'. 

Most operations in R are *vectorized*, meaning that by default they will happen component-wise when given a vector as an input. This is very useful to understand as it avoids a lot of unnecessary `for` loops!
```{r Vectorizing ex}
x <-  1:4

#Vectorized version of squaring 'x'
(y <- x^2)

#For loop version of squaring 'x'
  #Need something for the results of the for loop to be placed into
z <- vector(mode = mode(x), length = length(x))
z

for (i in seq_along(x)) {
  z[i] <- x[i]^2
}

#These end up with the same result
identical(y, z)
```

However, be unaware of vectorizing can also lead to traps, as (a) it can lead to unexpected results if you don't intend for this behaviour, and (b) it means that R will recycle vectors if they are not of the necessary length. It is good practice to check your R objects to make sure they are what you expect in terms of outputs, flavors, and length/dimensions!
```{r Recycling vectors}
(y <-  1:3)
(z <- 3:7)

y + z #R will warn you of uneven lengths here

(y <- 1:10)
(z <- 3:7)

y + z #But not here...I think because length(y) is a multiple of length(z)...but this is bad news!
```

#### Logical evaluations, if statments, and vectorization
The `|`, `&` and `==` evaluators are vectorized and apply to multiple values and should never be used in programming control flow and certain conditional statements (e.g. `if()`). 
Instead, you can use `||` and `&&` where the double operators returns `TRUE` for the first instance of `TRUE` it sees, and returns `FALSE` for the first instance of `FALSE` it sees. The double operators are "short circuits". Thus, while the single sign version compares at each element (i.e. is vectorized), while the double sign evaluates from left to right, examining only the first element of each vector, and evaluation stops once a result is determined.
In terms of output, the short form will result in a logical vector, while the double operators will return only one value. `if()` statements need a result of length 1 and will not work with a logical vector, hence why you should never use the single operators. However, a logical vector would be appropriate to an `ifelse()` statement.
```{r}
c(T, T, F) & c(T, F, F)
c(T, T, F) && c(T, F, F)
c(T, T, F) | c(T, F, F)
c(T, T, F) || c(T, F, F)

#example of short-circuiting (i.e. only evaluating as many terms as they need to get a result)
a
TRUE || a
FALSE || a
TRUE | a
FALSE & a
```
Alternatively, you can use `any()` or `all()` to collapse a vector (logical for `|` or `&`, or any type for `==` equivalencies) down to a single value. Another alternative for determining equivalency is to use `identical()`, although it is very strict and will not convert between different types (i.e. `identical(1L, 1)` returns `F`) and measures down to smallest possible value for integers and doubles, so use with caution! For comparing integers and doubles, `dyplyr::near()` may be a better option.

Don't forget that `TRUE` coerces to 1 and `FALSE` coerces to 0.

### Lists: 
Lists are like a vector but do not require that the elements are of the same flavor Data frames are a special case of a list where each element is an atomic vector, and all elements have the same length. Many functions will return lists to you and you will want to get at the list's innards. The components of a list also have names, and you can create or change names using the `names` function. 

Extracting elements from a list. See R for Data Science's [condiments](https://r4ds.had.co.nz/vectors.html#lists-of-condiments) section for great visual overview (also section 20.5.2 above the pepper shaker bit is quite useful too). Note that you have to use the double bracket approach, as opposed to the dollar sign, when the indexing object itself is an R object (eg "iWantThis" in the example below. Additionally, if you want more than one element, you must use vector-style indexing with **single** square brackets: this will always return another list (even if you request only 1 element). If you subset using double square brackets (for ONE element) you will be the flavor of that element.

Summary of how to extract using square brackets:

* x[1] returns the first element (eg data frame) in a list, still surrounded by a list wrapper

* x[[1]] returns only the first element (eg the data frame itself), without a list wrapper

* x[[1]][1] returns the first element of the data frame within our list, surrounded by a list wrapper

* x[[1]][[1]] returns the first element of the data frame within our list, without the list or data frame wrappers
```{r examples of how to extract elements from a list}
(a <- list(veg = c("cabbage", "eggplant"),
           tNum = c(pi, exp(1), sqrt(2)),
           myAim = TRUE,
           joeNum = 2:6))

# Explore this list
str(a)
length(a)
class(a)
mode(a) #Note this tells you the STORAGE mode, not the statistical mode!!!!

#Ways to get to a single element
a[[2]]    #index with a positive integer
a$myAim   #use dollar sign and element name
str(a$myAim) #we get myAim itself, a length of 1 logical vector
a[["tNum"]]  #index with a length 1 character vector
str(a[["tNum"]])  #we get tNum itself, a length 3 numeric vector
iWantThis <- "joeNum"  #indexing with a length 1 character object
a[[iWantThis]]  #we get joeNum itself, a length 5 integer vector
# a[[c("joeNum", "veg")]]  #This does NOT work!!! Can't get >1 elements!!! Can't knit with the operation either...

#Ways to get multiple elements
names(a)
a[c("tNum", "veg")]   #indexing by length 2 character vector
str(a[c("tNum", "veg")])  #Returns a list of length 2
str(a["veg"])   #returns a list of length 1
length(a["veg"])
length(a["veg"][[1]])  #contrast with the length of the veg vector itself
```

### Matrices and data flavors: 
A matrix is an alternative to data frames for storing rectangular data. Matrices are generalizations of atomic vectors, so all elements must be of the same class. Indexing is similar to that of a vector or a list, except that you can use `[i, j]`, where `i` = row and `j` = column. You can fill a matrix with a vector, by giving row and column numbers (and names) at creation. If the input vector is not long enough, R will recycle it.
```{r making matrices from vectors}
matrix(1:15, nrow = 5)
matrix("yo!", nrow = 3, ncol = 6)
matrix(c("yo!", "foo?"), nrow = 3, ncol = 6)
matrix(1:15, nrow = 5, byrow = TRUE)
matrix(1:15, nrow = 5,
       dimnames = list(paste0("row", 1:5),
                       paste0("col", 1:3)))
```

Note that if your elements are not of the same flavor, it may either convert everything to the lowest common flavor (usually character), or you will either get an error, or R can silently fail. This is why it is really important to understand the data flavors in your objects!!!
```{r }
multiDat <- data.frame(vec1 = 5:1,
                       vec2 = paste0("hi", 1:5))
str(multiDat)
(multiMat <- as.matrix(multiDat))
str(multiMat)  #Oh bad news - first element is now just 'NULL' instead of integers...
```

Table of atomic R object flavors (very handy!!!)
*Note the following will look great when rendered with pandoc but terrible in html, but is a good example of how to make a text table with centered values*
|:---------:|:-------------:|:---------:|:---------:|
| "flavor"  | type reported | mode()    | class()   |
|           | by typeof()   |           |           |
| character | character     | character | character |
| logical   | logical       | logical   | logical   |
| numeric   | integer       | numeric   | integer   |
|           | or double     |           | or double |
| factor    | integer       | numeric   | factor    |

```{r atomic R object flavors}
# Legible in any format
#  +-----------+---------------+-----------+-----------+
#  | "flavor"  | type reported | mode()    | class()   |
#  |           | by typeof()   |           |           |
#  +===========+===============+===========+===========+
#  | character | character     | character | character |
#  +-----------+---------------+-----------+-----------+
#  | logical   | logical       | logical   | logical   |
#  +-----------+---------------+-----------+-----------+
#  | numeric   | integer       | numeric   | integer   |
#  |           | or double     |           | or double |
#  +-----------+---------------+-----------+-----------+
#  | factor    | integer       | numeric   | factor    |
#  +-----------+---------------+-----------+-----------+
```

## Control structure and conditional statements: If, Ifelse, and Switch and Cut functions
Conditional statements are useful to controling how and when certain expressions are evaluted such as *if* `if(condition) { expression }`. If the condition of the *if* statment evaluates to FALSE, then the following expression is NOT executed and you will not get your result. If you want something done if the *if* condition evaluates to FALSE you can also add *else* `if(condition) { expr1 } else { expr2 }`, and/or *else if* `if(condition) { expr1 } else if(condition2) { expr2 } else { expr3 }` . Note that the `else` keyword **must** be on the same line as the closing bracket of the `if` statement!

A very important note with *if*, *else if*, and *else* statements is that R will execute through the code until it hits the **first condition** which evaluates to TRUE, then stops and does NOT execute the rest of the code. This is very important to keep in mind if you've made statements that are not mutuatally exclusive!!!

If you are getting confused about how your conditional statement is being evaluated, substitute your desired result with text and a print statement, so that you can see where in the conditional statement your results are being evaluated from (and if it is working as expected or not!!!)
```{r Trouble-shooting ifelse statements}
number <- 56

if(number < 10){
  if(number < 5){
    result <- "extra small"
  } else {
    result <- "small"
  }
  } else if(number < 100) {
    result <- "medium"
  } else {
    result <- "large"
  }
  print(result)
```

However, if you are writing a long series of chained *if* statements, you should instead consider using alternative functions.  

* `switch` evaluates selected code based on position or name. Format is: `switch(object, case = action)`
```{r switch ex}
basic_maths <- function(x, y, op) {
                  switch (op,
                    plus = x + y,
                    minus = x - y,
                    times = x * y,
                    divide = x / y,
                    stop("Unknown op!")
                  )
                }

basic_maths(x = 10, y = 5, op = "times")
```

* `cut` is useful to turn continuous variables into discrete categories. The argument 'breaks' can be supplied as a single number (which divides the data into n = break pieces of equal length) or a numeric vector of two or more unique cut points. The 'right' argument (default = T) specifies if intervals should be closed on the right and open on the left, or vice versa. There is also a 'labels' argument. Finally, note that there is also a `cut.Date` function for class 'Date' or 'POSIXt' objects. Note that 'cut.Date' does NOT work on date 'Interval' class.
```{r cut ex}
Z <- 1:100
table(cut(Z, breaks = 5, labels = F))
table(cut(Z, breaks = 5))
table(cut(Z, breaks = 5, right = F))


mydates <- seq(from = ymd("2016-09-01"), to = ymd("2018-08-30"), by = "week")
table(cut.Date(mydates, breaks = "month")) 
table(cut.Date(mydates, breaks = "month", labels = F)) 
```

* `stopifnot` is useful to check the validity of arguments. For instance, is your input numeric as expected? Are there NAs that need to be removed? Recall that `stopifnot` is evaluating if something is TRUE rather than you asserting what might be wrong.
```{r stopifnot ex}
# Simple example
mmm <- function(x){
  stopifnot(is.numeric(x))
  max(x) - min(x)
}

mmm(gapminder$country)

# More complex example
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}

wt_mean(1:6, 6:1, na.rm = "foo")
```

## Outline for writing your own function
When to write a function? The general rule of thumb seems to be if you've copied and pasted code 3 or more times. Yes, copy and paste is fast, but it is also WAY more prone to unnotice errors and will cause headaches down the line!!!

The process for writing a function follows the idea of building a minimum viable product: a limited-but-functioning thing is still very useful and easier to build into a more complex thing than just building the more complex thing straight from scratch. 

General steps are as follows:

1. Develop some working, interactive code. 

  i) This means that you start with a **simple** problem (or multiple simple problems). It is really important that you know what to expect for your output (e.g. data flavor, length, possible values)!!!
  
  ii) Then develop a working snippet of code that solves this problem(s). 
  
  iii) Test the snippet of code with test data, including some unexpected values (e.g. NAs, values outside of expected range, different flavors of data, different data types (vector vs data frame vs list)). This will give you an idea of when/where it might fail and if it fails, what you expect it to do. More on testing later.

2. Rewrite your snippet of code to use temporary variables. These temporary variables should be used where elements of code are repeated and/or when inputs are needed. Make sure to use **clear** names, not just a letter (in most cases) as it makes trouble-shooting down the line more difficult and readability for other users (i.e. future you!) much more difficult.

* Use temporary variables for the inputs; you can start with a very short, generic name to see where the inputs are (i.e. 'x') and then re-write the variable names for clarity

* Final variable names for inputs (aka argmuments) should be nouns

3. Re-write your code for clarity.

4. NOW turn your code snippet into a function by adding the `function(arguments) {}` part. Recall that the value of the last evaluated expression will be returned; however, if you call `return` within the function, that expression will be evaluated, the value returned, and the function will stop here (so **careful** if/where you use return calls!!!)

  i) Pick a descriptive name for your function. Function names should generally be verbs (although sometimes a noun may be more appropriate if your verb is very broad (e.g. get, compute, determine, calculate) For instance, 'get_mean()' probably should just be 'mean()'. Seperate words with an underscore (note that I've also seen R style guides say that words in objects and function should be seperated with a period...so just pick a convention and be consistent.) 
  
  ii) List the inputs (aka arguments) inside the function, for instance `function(x, y, z)`.
  
  iii) Place the working snippet of code inside the body of the functions (aka in the `{}` part)

5. Test that new function! 

  i) Use on some test data that mimics your intended data and where you (roughly) know what it should return.
  
  ii) Use on real data but *different* real data (e.g. other similar flavor variables in your dataset, an entirely different dataset with similar flavors of variables or data types like data frame or lists).
  
  iii) Use on weird stuff. Here you are trying to break your function. Use NAs, values outside of range, different flavors, different data types. 
  
  iv) Use automated tests! See the `testthat` package and the R packages book section on [testing](http://r-pkgs.had.co.nz/tests.html )
  
6. Based on results of step 5, ensure your function fails noisily and as soon as possible and insert informative error messages. See section below. Also note that you can skip this step if you are not intending to re-use your function; no sense in just making extra work for yourself.

## Making functions fail and writing informative error messages
Insert notes on stopifnot and informative error messages

Insert notes how to do validity checks on your functions to make sure they are doing what you think they are (see part 1 and part 2 examples)
And how to make them fail in expected ways


Notes on generalizing functions

```{r First function}
mmm <- function(x) {
  stopifnot(is.numeric(x))
  max(x) - min(x)
}
```

```{r Generalized function}
```

Reminder that function argument names can be anything but convention is to use meaningful names and if you are reusing functions or arugments in other functions, use the same names just to make it easier to remember (e.g. input_animals, input_inverts, input_plants vs. plants_input, inverts_input, plants_input)

Notes on how to return output from a function

Notes on default values

## For and While Loops
STILL NEED TO PUT NOTES IN HERE


## Functional programing with Purrr
LOOK AT DATA CAMP SLIDES?