---
title: "Functions in R"
author: "Michelle Coombe"
date: "13/02/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Load libraries and data
```{r}
library(tidyverse)
library(gapminder)
```
Notes in this document are primarily from the original [stat545](https://stat545.com/functions-part1.html), but also from the R for Data Science [ebook](https://r4ds.had.co.nz/functions.html), and the newer [ubcstat545](https://stat545guidebook.netlify.com/functional-programming-in-r-part-i.html).

## Write your own Functions, part 1
Insert notes on stopifnot and informative error messages

Insert notes how to do validity checks on your functions to make sure they are doing what you think they are (see part 1 and part 2 examples)
And how to make them fail in expected ways

## Write your own Functions, part 2
#### Tangent
You can get info out of (?some) plot objects without plotting it
For example, to get the 5 number summary from a boxplot (numbers between 0 and 100 rather than 0 and 1)
```{r boxplot summary}
boxplot(gapminder$lifeExp, plot = F)$stats
```
Cool!!! Good to remember that making the graph is actually a (desired) side effect of making the boxplot object

Notes on generalizing functions

```{r First function}
mmm <- function(x) {
  stopifnot(is.numeric(x))
  max(x) - min(x)
}
```

```{r Generalized function}
```

Reminder that function argument names can be anything but convention is to use meaningful names and if you are reusing functions or arugments in other functions, use the same names just to make it easier to remember

Notes on how to return output from a function

Notes on default values

## Write your own Functions, part 3